### 2015-04-18 implementation-specs.tex
GLL edits

\begin{figure}
	\subfigure[Spec for {\tt mpileaks}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
		\label{fig:specs-mpileaks}
	}
	\subfigure[{\tt mpileaks@2.3}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-version}
		\label{fig:specs-mpileaks-version}
	}
	\subfigure[{\tt mpileaks@2.3 \^{}callpath@1.0+debug \^{}libelf@0.8.11}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
		\label{fig:specs-mpileaks-abstract}
	}
	\subfigure[Spec (b), concretized.]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
		\label{fig:specs-mpileaks-concrete}
	}
	\caption{
		Specs for the {\tt mpileaks} package.
		\label{fig:specs}
	}
\end{figure}

@@

We begin with an example.%Let us explain our syntax with an example.
Consider the case where a user wants to install the {\tt mpileaks} package, but knows
nothing about its structure.  To install the package, the user writes:
%
\begin{minted}[fontsize=\scriptsize]{bash}
$ spack install mpileaks
\end{minted}
%
Here, {\tt mpileaks} is the simplest possible spec---a single identifier.
Spack parses it and converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
Note that even though the spec contains no dependency information, it is still 
converted to a full DAG, based on the directives supplied in package files. Since there
are no constraints on the nodes, Spack has considerable leeway for how to build the package,
and we say that the package is {\it unconstrained}.
%
Now, suppose the user wants a specific version of {\tt mpileaks}.  This can be requested
with a version constraint after the package name:
%
\begin{minted}[fontsize=\scriptsize]{bash}
$ spack install mpileaks@2.3
\end{minted}
%
We see from Figure~\ref{fig:specs-mpileaks-version} that the specific version constraint is
placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
If the user does not need a specific version of {\tt mpileaks}, but does require
particular minimu version, then the user could use {\it version range} syntax.
and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save build time if
Spack already has a version installed that satisfies the constraint -- Spack will just take
the already built instead of building a new one.

Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of spec syntax:
%
\begin{minted}[fontsize=\scriptsize]{bash}
$ spack install mpileaks@2.3 ^callpath@1.0+debug ^libelf@0.8.11
\end{minted}
%
The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
we now see that there are now version constraints on {\tt callpath} and {\tt libelf},
and the user has asked for the debug variant of the {\tt callpath} library.

Recall that Spack guarantees there will be only a single version of any package in 
the spec DAG.  Therefore, within the same DAG, each dependency can be uniquely identified by 
only its package name.  The user does not have to think about DAG connectivity to add
constraints.  She need only know that a package depends, somehow on {\tt callpath}. 
For the same reason, the constraint order does not matter; dependency constraints
can appear in arbitrary order.

@@

{\bf Variants.}
To handle build options like compiler flags or optional components, specs can
have named flags, or {\it variants}.  Variants are associated with the package,
so the {\tt mpileaks} package implementor must check the spec and handle the cases
where debug is enabled ({\tt +debug}) and disabled (with {\tt -debug}
or {\tt ~debug}).  The names simplify the versioning and prevent
Spack's configuration space from becoming too fine-grained.
It would violate our goal of conciseness, for example, to include detailed 
compiler flags in the spec syntax, but known sets of flags can simply be named.

@@

So far, we have shown examples of specs being used to request constraints from the
command line, when {\tt spack install} is invoked.  However, the user is not the only
source of constraints.  Applications may require specific versions of dependencies,
and it is often desirable to write constraints like this into a package file.  For
example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
At first glance, the {\tt depends\_on()} directive in Figure~\ref{fig:mpileaks} look
like they take simple package names.  However, a package name is also a spec, and
the same constraint syntax usable from the command line can be applied inside directives.
So, the ROSE compiler can simply write:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('boost@1.54.0')
\end{minted}
%
This constraint will be incorporated into the initial DAG node generated from
the ROSE package.

Spack's {\tt patch} directive also accepts spec syntax as a predicates in an
optional {\tt when} parameter.  For example, these concise directives in the 
Python package ensure that specific patches are applied to the Python source
code when it is being built on Blue Gene/Q, with the appropriate compiler:
%
\begin{minted}[fontsize=\scriptsize]{python}
    patch('python-bgq-xlc.patch',   when='=bgq%xlc')
    patch('python-bgq-clang.patch', when='=bgq%clang')
\end{minted}
%
Constraints in the {\tt when} clause are matched against the Python package spec.
Outside of directives, constraints can be used directly
with the {\tt spec} object in the {\tt install} method:
%
\begin{minted}[fontsize=\scriptsize]{python}
  def install(self, spec, prefix):
      if spec.satisfies('%gcc'):
          # Handle gcc
      elif spec.satisfies('%xlc'):
          # Handle XL compilers
      ...
\end{minted}
%

@@

For cases like this, Spack provides functionality that allows Python functions to have
multiple definitions, with some specialized for particular configurations of the package.
This allows us to have two separate implementations of {\tt install} or {\it any} method 
in a package class. Figure~\ref{fig:specialization} shows how this is used in the Dyninst
package.  The {\tt @when} directive is a Python decorator: a higher order function that 
takes a function definition as a parameter and returns a new function to put in its place.
Here, we replace the function with a callable multi-function dispatch object, and we 
integrate the predicate check into the function dispatch mechanism.  Here, the
{\tt when} condition is true when Dyninst is at version 8.1 or lower, and in those cases
the package will use the {\tt configure}-based build.  By default if no predicate matches,
{\tt install} will simply use the default CMake-based implementation. The simple {\tt @when}
annotation allows us to maintain our old build code alongside the new version without 
accumulating complex logic in a single {\tt install} function.

>>>>

\begin{figure}
	\begin{subfigure}{\linewidth}
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
		\caption{
			Spec for {\tt mpileaks}
			\label{fig:specs-mpileaks}
		}
	\end{subfigure}
%
	\begin{subfigure}{\linewidth}
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-version}
		\caption{
			{\tt mpileaks@2.3}
			\label{fig:specs-mpileaks-version}
		}
	\end{subfigure}
%
	\begin{subfigure}{\linewidth}
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
		\caption{
			{\tt mpileaks@2.3 \^{}callpath@1.0+debug \^{}libelf@0.8.11}
			\label{fig:specs-mpileaks-abstract}
		}
	\end{subfigure}
%
	\caption{
		Constraints applied to {\tt mpileaks} specs.
	}
\end{figure}

@@

We begin with a simple example.
Consider the case where a user wants to install the {\tt mpileaks} package, but knows
nothing about its structure.  To install the package, the user invokes the {\tt spack install}
command:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks
\end{minted}
%
Here, {\tt mpileaks} is the simplest possible spec---a single identifier.
Spack parses it and converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
Note that even though the spec contains no dependency information, it is still 
converted to a full DAG, based on the directives supplied in package files. Since there
are no constraints on the nodes, Spack has considerable leeway for how to build the package,
and we say that the package is {\it unconstrained}.
%
Now, suppose the user wants a specific version of {\tt mpileaks}.  This can be requested
with a version constraint after the package name:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks@2.3
\end{minted}
%
We see from Figure~\ref{fig:specs-mpileaks-version} that the specific version constraint is
placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
If the user does not need a specific version of {\tt mpileaks}, but does require
particular minimum version, then the user could use {\it version range} syntax.
and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save build time if
Spack already has a version installed that satisfies the constraint -- Spack will just use
the previously-built installation instead of building a new one.

Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of the spec syntax:
%
\begin{minted}[fontsize=\scriptsize]{bash}
    $ spack install mpileaks@2.3 ^callpath@1.0+debug ^libelf@0.8.11
\end{minted}
%
The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
we now see that there are version constraints on {\tt callpath} and {\tt libelf},
and that the user has asked for the debug variant of the {\tt callpath} library.

Recall that Spack guarantees there will be only a single version of any package in 
the spec DAG.  Therefore, within the same DAG, each dependency can be uniquely identified by 
only its package name.  The user does not have to think about DAG connectivity to add
constraints.  She need only know that a package depends, somehow, on {\tt callpath}. 
For the same reason, the constraint order does not matter; dependency constraints
can appear in arbitrary order.

@@

{\bf Variants.}
To handle build options like compiler flags or optional components, specs can
have named flags, or {\it variants}.  Variants are associated with the package,
so the {\tt mpileaks} package implementor must check the spec and handle the cases
where debug is enabled ({\tt +debug}) and disabled (with {\tt -debug}
or {\tt ~debug}).  The names simplify the versioning and prevent
Spack's configuration space from becoming too fine-grained.
It would violate our goal of conciseness, for example, to include detailed 
compiler flags in spec syntax, but known sets of flags can simply be named.

@@

So far, we have shown examples of specs being used to request constraints from the
command line, when {\tt spack install} is invoked.  However, the user is not the only
source of constraints.  Applications may require specific versions of dependencies,
and it is often desirable to write constraints like this into a package file.  For
example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
At first glance, the {\tt depends\_on()} directives in Figure~\ref{fig:mpileaks} look
like they take simple package names.  However, a package name is also a spec, and
the same constraint syntax usable from the command line can be applied inside directives.
So, the ROSE compiler can simply write:
%
\begin{minted}[fontsize=\scriptsize]{python}
    depends_on('boost@1.54.0')
\end{minted}
%
This constraint will be incorporated into the initial DAG node generated from
the ROSE package.

Spack's {\tt patch} directive also accepts spec syntax as a predicate in an
optional {\tt when} parameter.  For example, these concise directives in the 
Python package ensure that specific patches are applied to the Python source
code when it is being built on Blue Gene/Q, with the appropriate compiler:
%
\begin{minted}[fontsize=\scriptsize]{python}
    patch('python-bgq-xlc.patch',   when='=bgq%xlc')
    patch('python-bgq-clang.patch', when='=bgq%clang')
\end{minted}
%
Constraints in the {\tt when} clause are matched against the Python package spec.
Outside of directives, constraints can be used directly
with the {\tt spec} object in the {\tt install} method:
%
\begin{minted}[fontsize=\scriptsize]{python}
  def install(self, spec, prefix):
      if spec.satisfies('%gcc'):
          # Handle gcc
      elif spec.satisfies('%xlc'):
          # Handle XL compilers
      ...
\end{minted}
%

@@

For cases like this, Spack provides functionality that allows Python functions to have
multiple definitions, with some specialized for particular configurations of the package.
This allows us to have two separate implementations of {\tt install} or {\it any} method 
in a package class. Figure~\ref{fig:specialization} shows how this is used in the Dyninst
package.  The {\tt @when} directive is a Python decorator: a higher order function that 
takes a function definition as a parameter and returns a new function to put in its place.
We replace the function with a callable multi-function dispatch object, and we 
integrate the predicate check into the function dispatch mechanism.  Here, the
{\tt when} condition is true when Dyninst is at version 8.1 or lower, and in those cases
the package will use the {\tt configure}-based build.  By default if no predicate matches,
{\tt install} will use the default CMake-based implementation. The simple {\tt @when}
annotation allows us to maintain our old build code alongside the new version without 
accumulating complex logic in a single {\tt install} function.
